The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free e-learning program CrypTool\\ (10th edition -- distributed with CrypTool version 1.4.30)} \vspace{60pt}

### 2009-08-10 script/de/script-de.tex
new float cryptoprocedure for crypto algorithms and procedures (chapter 5); fixed links to un-numbered chapters with phantomsection
resolved more pdfTeX warnings by adding a chapter specific prefix "xy:" to all cite and bibitem in chapters with bibliography
fixed overfull hbox
CT script, cover with smaller size; many small corrections; consistency for all chapters like bibliography everywhere in PDF navigation and link moves to correct page (de+en)

\documentclass[a4paper,11pt,oneside]{book}
\usepackage[latin1]{inputenc} % Umlaute
% \usepackage[dvips]{graphics}
% \usepackage{graphicx}

@@

% Formatierung der Fußnoten
\usepackage[flushmargin]{footmisc}
\usepackage{fullpage}
\usepackage{url}
% Literaturverzeichnis kapitelweise
\usepackage[labelstoglobalaux]{bibunits} % ohne labelstoglobalaux scheint man thebibilograpy und bibtex nicht mischen zu können
\usepackage{float}
\usepackage[colorlinks=true,linkcolor=black,citecolor=black,urlcolor=black,bookmarksnumbered=true,pdfpagelabels,hypertexnames=false]{hyperref}
%To reduce the number of hyperlink related 'pdfTeX warning's we use options 
%pdfpagelabels and hypertexnames=false and place hyperref after float.
%Option debug helps finding problems.

\floatstyle{ruled}
\newfloat{sagecode}{!h}{loc}[chapter]
\floatname{sagecode}{Sage-Beispiel}

@@

% \setlength{\textwidth}{460 pt}
% \hoffset-0.7in
\newtheorem{definition}{Definition}[section]
\newtheorem{satz}{Satz}[section]
\newenvironment{Beweis}[1]
  {\textbf{Beweis #1} \\}
  {\hfill$\Box$ \\}
\newenvironment{example}[1]{\noindent\textbf{Beispiel#1}}{}
\newenvironment{remark}[1]{\noindent\textbf{Bemerkung#1}}{}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{\vspace{25pt}
	{\Huge Das CrypTool-Skript: \\
	\vspace{5pt}
	Kryptographie, Mathematik und mehr} \\
        \vspace{20pt}
        {\LARGE Hintergrundmaterial und Zusatzinformationen\\
         zum freien E-Learning-Programm CrypTool} \\
	\vspace{15pt}
        {\large (10. Auflage -- veröffentlicht mit CrypTool-Version 1.4.30)}
        \vspace{180pt}
}

@@

% Vergleichbar zur Nutzung von Texten aus der Wikipedia (=GNU Free Documentation Licence, GFDL)
\vspace{200pt}
\noindent Dies ist ein freies Dokument, d.h. Inhalte des Dokuments können kopiert und
verbreitet werden, auch zu kommerziellen Zwecken. \\
Im Gegenzug sind Autor, Titel und die CrypTool-Webseite (www.cryptool.org) zu
nennen. Selbstverständlich darf aus dem CrypTool-Skript, genau wie aus jedem
anderen Werk auch, zitiert werden. \\
Dieses Dokument unterliegt der GNU-Lizenz für freie Dokumentation
\hyperlink{appendix-GNU-fdl}{(siehe Anhang)}.  % HIER erh. den selbstbest. Text als link.
% (siehe \ref{s:appendix-GNU-fdl}).  % HIER erh. man eine Kap.Nr. "8.6".

\begin{quote}
    Copyright \copyright{} 1998--2009 Bernhard Esslinger and the
    CrypTool Development Team. Permission is granted to copy,
    distribute and/or modify this document under the terms of the GNU
    Free Documentation License, Version 1.3 or any later version
    published by the Free Software Foundation; with no Invariant
    Sections, no Front-Cover Texts, and no Back-Cover Texts. A copy of
    the license is included in the section entitled ``GNU Free
    Documentation License''.
\end{quote}

@@

%\appendix \addcontentsline{toc}{section}{Appendix} %be_2005 raus
\begin{appendix}
  \newpage
% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
%\newpage  be_2006 Nötig ?
\hypertarget{appendix-start}{}\label{s:appendix-start}
\chapter{Anhang}
    \begin{itemize}
      \item[1] \hyperlink{appendix-menutree}{CrypTool-Menübaum}
      \item[2] \hyperlink{appendix-authors}{Autoren des CrypTool-Skripts}
      \item[3] \hyperlink{appendix-movies}{Filme und belletristische
                          Literatur mit Bezug zur Kryptographie, Bücher
                          mit Sammlungen einfacher Verfahren für Kinder}
      \item[4] \hyperlink{appendix-Learn-NT}
                         {Lernprogramm Elementare Zahlentheorie}
      \item[5] \hyperlink{appendix-using-sage}
                         {Kurzeinführung in das Computeralgebrasystem Sage}
    \end{itemize}
  \input{chapters/menus.tex}
  \input{chapters/authors.tex}
  \input{chapters/what-is-sage.tex}
\end{appendix}

\input{chapters/gnu-fdl-1.3.tex}
% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\begingroup
% remove extra vertical space from \listoffigure and \listoftables so that 
% it look similar to \listof{sagecode}
\setlength{\parskip}{0pt}
\clearpage
\listoffigures
\addcontentsline{toc}{chapter}{\listfigurename}

\clearpage
\listoftables
\addcontentsline{toc}{chapter}{\listtablename}

\clearpage
\listof{sagecode}{Verzeichnis der Sage-Programmbeispiele}
\addcontentsline{toc}{chapter}{Sage-Programmbeispiele}
\endgroup

%\addcontentsline{toc}{section}{Index} %moved to style.ist

\printindex

>>>>

\documentclass[a4paper,11pt,oneside]{book}
%\overfullrule=5pt % uncomment to mark overfull boxes
\usepackage[latin1]{inputenc} % Umlaute
% \usepackage[dvips]{graphics}
% \usepackage{graphicx}

@@

% Formatierung der Fußnoten
\usepackage[flushmargin]{footmisc}
\usepackage{fullpage}
\usepackage[hyphens]{url} % options "hyphens" enables linebreaks in \url{}
% Literaturverzeichnis kapitelweise
\usepackage[labelstoglobalaux]{bibunits} % ohne labelstoglobalaux scheint man thebibilograpy und bibtex nicht mischen zu können
\usepackage{float}
\usepackage[colorlinks=true,linkcolor=black,citecolor=black,urlcolor=black,bookmarksnumbered=true,pdfpagelabels,hypertexnames=false]{hyperref}
%To reduce the number of hyperlink related 'pdfTeX warning's we use options 
%pdfpagelabels and hypertexnames=false and place hyperref after float.
%Option debug helps finding problems.

\floatstyle{ruled}
\newfloat{sagecode}{!h}{loc}[chapter]
\floatname{sagecode}{Sage-Beispiel}
\newfloat{cryptoprocedure}{!h}{lop}[chapter]
\floatname{cryptoprocedure}{Krypto-Verfahren}

@@

% \setlength{\textwidth}{460 pt}
% \hoffset-0.7in
\newtheorem{definition}{Definition}[section]
\newtheorem{satz}{Satz}[section]
\newenvironment{Beweis}[1]{\noindent\textbf{Beweis #1} \\}{\hfill$\Box$\par}
\newenvironment{example}[1]{\noindent\textbf{Beispiel#1}}{}
\newenvironment{remark}[1]{\noindent\textbf{Bemerkung#1}}{}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{\vspace{25pt}
	{\Huge Das CrypTool-Skript: \\
	\vspace{5pt}
	Kryptographie, Mathematik und mehr} \\
        \vspace{20pt}
        {\LARGE Hintergrundmaterial und Zusatzinformationen\\
         zum freien E-Learning-Programm CrypTool\\
	 (mit Code-Beispielen zur Zahlentheorie, geschrieben in Sage)} \\
	\vspace{15pt}
        {\large (10. Auflage -- veröffentlicht mit CrypTool-Version 1.4.30)}
        \vspace{180pt}
}

@@

% Vergleichbar zur Nutzung von Texten aus der Wikipedia (=GNU Free Documentation Licence, GFDL)
\begin{quote}
Dies ist ein freies Dokument, d.h. Inhalte des Dokuments können kopiert und
verbreitet werden, auch zu kommerziellen Zwecken. \\
Im Gegenzug sind Autor, Titel und die CrypTool-Webseite (www.cryptool.org) zu
nennen. Selbstverständlich darf aus dem CrypTool-Skript, genau wie aus jedem
anderen Werk auch, zitiert werden. \\
Dieses Dokument unterliegt der GNU-Lizenz für freie Dokumentation.

    Copyright \copyright{} 1998--2009 Bernhard Esslinger and the
    CrypTool Development Team. Permission is granted to copy,
    distribute and/or modify this document under the terms of the GNU
    Free Documentation License, Version 1.3 or any later version
    published by the Free Software Foundation; with no Invariant
    Sections, no Front-Cover Texts, and no Back-Cover Texts. A copy of
    the license is included in the section entitled
    \hyperlink{appendix-GNU-fdl}{``GNU Free Documentation License''}.
\end{quote}

\vspace{450pt}
Quelle Coverfoto: \url{www.photocase.com}, Andre Günther

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\begin{appendix}
\newpage
\hypertarget{appendix-start}{}\label{s:appendix-start}
\chapter{Anhang}
    \begin{itemize}
      \item[1] \hyperlink{appendix-menutree}{CrypTool-Menübaum}
      \item[2] \hyperlink{appendix-authors}{Autoren des CrypTool-Skripts}
      \item[3] \hyperlink{appendix-movies}{Filme und belletristische
                          Literatur mit Bezug zur Kryptographie, Bücher
                          mit Sammlungen einfacher Verfahren für Kinder}
      \item[4] \hyperlink{appendix-Learn-NT}
                         {Lernprogramm Elementare Zahlentheorie}
      \item[5] \hyperlink{appendix-using-sage}
                         {Kurzeinführung in das Computeralgebrasystem Sage}
    \end{itemize}
  \input{chapters/menus.tex}
  \input{chapters/authors.tex}
  \input{chapters/what-is-sage.tex}
\end{appendix}

\input{chapters/gnu-fdl-1.3.tex}

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\begingroup
% remove extra vertical space from \listoffigure and \listoftables so that 
% it look similar to \listof{sagecode}
\setlength{\parskip}{0pt}
\clearpage\phantomsection
\addcontentsline{toc}{chapter}{\listfigurename}
\listoffigures

\clearpage\phantomsection
\addcontentsline{toc}{chapter}{\listtablename}
\listoftables

\clearpage\phantomsection
\addcontentsline{toc}{chapter}{Verzeichnis der Krypto-Verfahren}
\listof{cryptoprocedure}{Verzeichnis der Krypto-Verfahren}
\endgroup

\clearpage\phantomsection
\addcontentsline{toc}{chapter}{Verzeichnis der Sage-Programmbeispiele}
\listof{sagecode}{Verzeichnis der Sage-Programmbeispiele}

\clearpage\phantomsection
\addcontentsline{toc}{chapter}{Index}
\printindex

### 2009-08-10 script/en/chapters/aboutcryptool.tex
new float cryptoprocedure for crypto algorithms and procedures (chapter 5); fixed links to un-numbered chapters with phantomsection
CT script, cover with smaller size; many small corrections; consistency for all chapters like bibliography everywhere in PDF navigation and link moves to correct page (de+en)

% --------------------------------------------------------------------------
\chapter*{Overview about the Content of the CrypTool Script}  
\addcontentsline{toc}{chapter}{Overview}

\parskip 4pt
%\vskip +12 pt
{
\noindent In this CrypTool\index{CrypTool} script you will find predominantly
mathematically oriented information on using cryptographic procedures. There is
also sample code in the CAS (computer algebra system) Sage for some of the
procedures.
The main chapters have been written by various {\bf authors}
(see appendix \ref{s:appendix-authors}) %\hyperlink{appendix-authors}{authors}
and are therefore independent from one another. At the end of most chapters
you will find references and web links.

>>>>

% --------------------------------------------------------------------------
\clearpage\phantomsection
\addcontentsline{toc}{chapter}{Overview}
\chapter*{Overview about the Content of the CrypTool Script}

\parskip 4pt
%\vskip +12 pt
\noindent In this {\em scipt accpmpaying the program CrypTool} \index{CrypTool}
you will find predominantly mathematically oriented information on using
cryptographic procedures. Also included are many sample code pieces in the CAS
(computer algebra system) Sage\index{Sage} for some of the procedures.
The main chapters have been written by various {\bf authors}
(see appendix \ref{s:appendix-authors}) %\hyperlink{appendix-authors}{authors}
and are therefore independent from one another. At the end of most chapters
you will find references and web links.

### 2009-08-10 script/en/chapters/introduction.tex
CT script, cover with smaller size; many small corrections; consistency for all chapters like bibliography everywhere in PDF navigation and link moves to correct page (de+en)

%This edition of the script adds some topics, but mainly updates areas (e.g. the
%summaries of topical research areas):
This edition completely updated the TeX sources of the document and of course
the content of the script was corrected, amended and updated with some topics, e.g.:
\vspace{-7pt}
\begin{itemize}
  \item the search for the largest prime numbers 
        (chap. \ref{search_for_very_big_primes}), 
  \item progress in cryptanalysis of hash algorithms 
        (chap. \ref{NeueAES-Analyse}) and
%  \item progress in cryptanalysis of hash algorithms 
%        (chap. \ref{collision-attacks-against-sha-1}) and
%  \item progress in ideas for new crypto methods (RSA successor) 
%        (chap. \ref{xxxxxxxxxBrute-force-gegen-Symmetr})\index{xxxxxxxxxxxxxxxx} and
  \item the list of movies or novels, in which cryptography or number theory 
        played major role (see appendix \ref{s:appendix-movies});
        and where primes are used as hangers  
        (see curiouses in \ref{HT-Quaint-curious-Primes-usage}).
\end{itemize}

@@

\vspace{8pt}
The first time the document was delivered with CrypTool\index{CrypTool} 
was in version 1.2.01. Since then it has been expanded and revised in almost
every new version of CrypTool.

@@

I hope that many readers have fun with this script and that they get 
out of it more interest and greater understanding of this modern but 
also very ancient topic.
\\
\\
% [1.5\baselineskip]
% \enlargethispage*{2\baselineskip}
% \nopagebreak
Bernhard Esslinger
\\
\\
% [\baselineskip]
Frankfurt (Germany), July 2009
\vspace{-12pt}

@@

This document is delivered together with the program CrypTool\index{CrypTool}.
\par \vskip + 3pt

@@

The authors
% \hyperlink{appendix-authors}{(authors)}
% in \ref{s:appendix-authors}
% \hypertarget{appendix-authors}{}\label{s:appendix-authors}
have attempted to describe cryptography for a broad 
audience -- without being mathematically incorrect. We believe that this
didactic pretension is the best way to promote the awareness for IT
security and the readiness to use standardized modern cryptography.
\par \vskip + 15pt

\textbf{The program CrypTool\index{CrypTool}}

CrypTool\index{CrypTool} is an educational program with a comprehensive online
help enabling you to use and analyse cryptographic procedures within a
unified graphical user interface.\par \vskip + 3pt

CrypTool\index{CrypTool} is currently been used for 
training in companies and teaching at school and universities worldwide, and
several universities are helping to further develop the project.
\par \vskip + 15pt

\textbf{Acknowledgment}

>>>>

%This edition of the script adds some topics, but mainly updates areas (e.g. the
%summaries of topical research areas):
This edition completely updated the TeX sources of the document, and of course
the content of the script was corrected, amended and updated with some topics, e.g.:
\vspace{-7pt}
\begin{itemize}
  \item the search for the largest prime numbers 
        (chap. \ref{search_for_very_big_primes}), 
  \item progress in cryptanalysis of hash algorithms 
        (chap. \ref{NeueAES-Analyse}) and
%  \item progress in cryptanalysis of hash algorithms 
%        (chap. \ref{collision-attacks-against-sha-1}) and
%  \item progress in ideas for new crypto methods (RSA successor) 
%        (chap. \ref{xxxxxxxxxBrute-force-gegen-Symmetr})\index{xxxxxxxxxxxxxxxx} and
  \item the list of movies or novels, in which cryptography or number theory 
        played major role (see appendix \ref{s:appendix-movies});
        and where primes are used as hangers  
        (see curiouses in \ref{HT-Quaint-curious-Primes-usage}).
\end{itemize}

@@

The first time the document was delivered with CrypTool\index{CrypTool} 
was in version 1.2.01. Since then it has been expanded and revised in almost
every new version of CrypTool.

@@

I hope that many readers have fun with this script and that they get 
out of it more interest and greater understanding of this modern but 
also very ancient topic.
\\
\\
% [1.5\baselineskip]
% \enlargethispage*{2\baselineskip}
% \nopagebreak
Bernhard Esslinger
\\
\\
% [\baselineskip]
Frankfurt (Germany), August 2009

@@

This document is delivered together with the open-source program CrypTool\index{CrypTool}.

@@

The \hyperlink{appendix-authors}{authors}
have attempted to describe cryptography for a broad 
audience -- without being mathematically incorrect. We believe that this
didactic pretension is the best way to promote the awareness for IT
security and the readiness to use standardized modern cryptography.
\par \vskip + 15pt

\noindent \textbf{The program CrypTool\index{CrypTool}}

CrypTool\index{CrypTool} is an educational program with a comprehensive online
help enabling you to use and analyse cryptographic procedures within a
unified graphical user interface.

CrypTool\index{CrypTool} is used world-wide for 
training in companies and teaching at schools and universities worldwide, and
several universities are helping to further develop the project.
\par \vskip + 15pt

\noindent \textbf{Acknowledgment}

### 2009-08-10 script/en/chapters/menus.tex
CT script: Spellcheck in order to have consistent American English
added argument to example and remark definitions
CT script, cover with smaller size; many small corrections; consistency for all chapters like bibliography everywhere in PDF navigation and link moves to correct page (de+en)

This appendix contains at the following page the complete menu tree of
CrypTool\index{CrypTool} version 1.4.30.

@@

%\nobreak
\clearpage
\begin{figure}[hb]
\begin{center}
\vspace{-30pt}
%\frame{  %TeX macht einen Rehmen (siehe viewport) drumherum -- gut zum Testen
\includegraphics[scale=0.25, angle=270, viewport=200 30 2660 1430]
                {figures/cryptool-menu-en}
%viewport=rand-links? rand-unten breite hoehe? [bezogen auf querformat]
%}
\caption{Complete overview of the menu tree of CrypTool 1.4.30} 
\label{menuoverview}
\end{center}
\end{figure}
\clearpage

>>>>

This appendix contains at the following page the complete menu tree of
CrypTool\index{CrypTool} version 1.4.30\footnote{%
  The future versions CrypTool 2 and JCrypTool of the CrypTool project are
  currently there only as (stable) beta versions. If the according release
  versions are available, we will add the appropriate menus and menu trees.\\
}.

@@

%\nobreak
\clearpage
\begin{figure}[hb]
\begin{center}
\vspace{-30pt}
%\frame{  %TeX macht einen Rahmen (siehe viewport) drumherum -- gut zum Testen
%\includegraphics[scale=0.25, angle=270, viewport=200 30 2660 1430]
\includegraphics[scale=0.25, angle=270]
                {figures/cryptool-menu-en}
%viewport=rand-links? rand-unten breite hoehe? [bezogen auf querformat]
%}
\caption{Complete overview of the menu tree of CrypTool 1.4.30} 
\label{menuoverview}
\end{center}
\end{figure}
\clearpage

### 2009-08-10 script/en/chapters/moderncryptography.tex
CT script, cover with smaller size; many small corrections; consistency for all chapters like bibliography everywhere in PDF navigation and link moves to correct page (de+en)

%--------------------------------------------------------------------
\subsection{ElGamal public key encryption procedure in $ \Z_p^\ast$}
\index{ElGamal!public key}
By varying the Diffie-Hellman key agreement protocol \index{Diffie-Hellman} 
\index{Encryption!ElGamal public key} slightly, you can obtain an 
asymmetric encryption algorithm, crypto procedure~\ref{elgamalproc}. 
This observation was made by Taher ElGamal.
\begin{cryptoprocedure}
Let $ p $ be a prime number such that the discrete logarithm in $ \Z_p $ is 
difficult to compute. Let $ \alpha \in \Z_p^\ast $ be a primitive element. Let $a \in \N$ and $ 
\beta = \alpha^{a}  \mod p. $\\
\textbf{Public key:}
\[ p,\alpha,\beta. \]
\textbf{Private key:}
\[a. \]
Let $ k \in \Z_{p-1} $ be a random number and $ x \in \Z_p^{\ast} $ the 
plaintext. \\
\textbf{Encryption:}
\[ e_T(x,k)=(y_1,y_2), \]
where
\[ y_1=\alpha^k \mod p\]
and
\[ y_2 = x\beta^k \mod p.\]
\textbf{Decryption:}
\[ d_T(y_1,y_2)= y_2 (y_1^{a})^{-1} \mod p \]
\caption{ElGamal (based on the factorization problem)}
\label{elgamalproc}
\end{cryptoprocedure}

@@

The discrete logarithm can be generalised in any number of finite \index{Group} 
groups $ (G, \circ) $. The following provides several properties of $ G, $ that 
make the discrete logarithm problem difficult. \\
\index{Exponential function!calculation}
\paragraph{Calculating the discrete exponential function}
Let $ G $ be a group with the operation $ \circ $ and $ g \in G. $ The 
(discrete) exponential function  base $ g $ is defined as
\[ e_g : k \longmapsto g^k, \quad {\rm for~all~} k \in {\rm N}. \]
where
 \[ \ g^{k}:=\underbrace{g \circ \ldots \circ g}_{k {\rm ~times}}.\]
The exponential function is easy to calculate:
\vskip +20 pt \noindent
{\bf Lemma.}

{\em
  The power $ g^k $ can be calculated in at most $ 2 \log_2 k $ group 
operations.
}
\vskip +10 pt

\begin{Proof}{}
Let $ k=2^n + k_{n-1} 2^{n-1} + \cdots + k_1 2 + k_0 $ be the binary 
representation of $k. $ Then $ n \leq \log_2 (k), $  because $ 2^n \leq k < 
2^{n+1}. $ $ k $ can be written in the form $ k=2k' + k_0 $ with $ k'= 2^{n-1} + 
k_{n-1} 2^{n-2} + \cdots + k_1 $. Thus 
\[ g^k = g^{2k'+k_0}= (g^{k'})^2 g^{k_0} .\]
We therefore obtain $ g^k $ from $ g^{k'} $ by squaring and then multiplying by 
$ g $. The claim is thus proved by induction to $ n. $
\end{Proof}

\vskip +10 pt
{\bf Problem of the discrete logarithm}\index{Logarithm problem!discrete}
\vskip +10 pt
\begin{center}
\fbox{\parbox{12cm}{        
Let $ G $ by a finite group with the operation $ \circ. $ Let $ \alpha \in G $ 
and $ \beta \in H=\{ \alpha^{i}: i\geq 0\}. $ \\
We need to find a unique $ a \in {\rm N} $ with $ 0 \leq a \leq |H| -1 $ and $ \beta 
= \alpha^{a}. $ \\       
We define $ a $ as $ \log_\alpha (\beta). $
}}
\end{center}

@@

If $ |G| $ contains a ``dominant'' prime factor $ p ,$ then the 
complexity \index{Complexity} of the logarithm problem is approximately
\[ O(\sqrt{p}). \]
Therefore, if the logarithm problem is to be made difficult, the order of the 
group used $ G $ should have a large prime factor. In particular, if the 
discrete exponential function in the group $ \Z_p^{\ast} $ is to be a one way 
function, then $ p -1 $ must be a large prime factor. In this case a
generatlized ElGamal procedure can be defined (crypto
procedure~\ref{genelgamalproc}).
\begin{cryptoprocedure}
Let $ G $ be a finite group with operation $  \circ, $ and let $ \alpha \in G, $ 
so that the discrete logarithm in $ H =\{ \alpha^{i}: i \geq 0 \} $ is 
difficult, Let $ a $ with $   0 \leq a \leq |H| -1 $ and let $ \beta = 
\alpha^{a}. $\\
\textbf{Public key:}
\[ \alpha,\beta. \]
\textbf{Private key:}
\[a. \]
Let $ k \in \Z_{|H|} $ be a random number and $ x \in G $ be a plaintext. \\
\textbf{Encryption:}
\[ e_T(x,k)=(y_1,y_2), \]
where
\[ y_1=\alpha^k \]
and
\[ y_2 = x\circ \beta^k .\]
\textbf{Decryption:}
\[ d_T(y_1,y_2)= y_2\circ (y_1^{a})^{-1}  \]
\caption{Generalized ElGamal (based on the factorization problem)}
\label{genelgamalproc}
\end{cryptoprocedure}

\hyperlink{ellcurve}{Elliptic curves} {} provide useful groups for public key 
encryption procedures.

@@

%--------------------------------------------------------------------
\section*{Web links}
\addcontentsline{toc}{section}{Web links}
\begin{enumerate}
   \item \href{http://www.geocities.com/st_busygin/clipat.html}
              {http://www.geocities.com/st\_busygin/clipat.html }
\end{enumerate}

>>>>

%--------------------------------------------------------------------
\subsection{ElGamal public key encryption procedure in $ \Z_p^\ast$}
\index{ElGamal!public key}
\index{Encryption!ElGamal public key}

By varying the Diffie-Hellman key agreement protocol\index{Diffie-Hellman} 
slightly, you can obtain an asymmetric encryption algorithm, crypto
procedure~\ref{elgamalproc}. This observation was made by Taher ElGamal.
\begin{cryptoprocedure}
Let $ p $ be a prime number such that the discrete logarithm in $ \Z_p $ is 
difficult to compute. Let $ \alpha \in \Z_p^\ast $ be a primitive element. Let $a \in \N$ and $ 
\beta = \alpha^{a}  \mod p. $\\
\textbf{Public key:}
\[ p,\alpha,\beta. \]
\textbf{Private key:}
\[a. \]
Let $ k \in \Z_{p-1} $ be a random number and $ x \in \Z_p^{\ast} $ the 
plaintext. \\
\textbf{Encryption:}
\[ e_T(x,k)=(y_1,y_2), \]
where
\[ y_1=\alpha^k \mod p\]
and
\[ y_2 = x\beta^k \mod p.\]
\textbf{Decryption:}
\[ d_T(y_1,y_2)= y_2 (y_1^{a})^{-1} \mod p \]
\caption{ElGamal (based on the discrete logarithm problem)}
\label{elgamalproc}
\end{cryptoprocedure}

@@

The discrete logarithm can be generalised in any number of finite \index{Group} 
groups $ (G, \circ) $. The following provides several properties of $ G, $ that 
make the discrete logarithm problem difficult. \\
\index{Exponential function!calculation}
\paragraph{Calculating the discrete exponential function}
Let $ G $ be a group with the operation $ \circ $ and $ g \in G. $ The 
(discrete) exponential function  base $ g $ is defined as
\[ e_g : k \longmapsto g^k, \quad {\rm for~all~} k \in {\rm N}. \]
where
 \[ \ g^{k}:=\underbrace{g \circ \ldots \circ g}_{k {\rm ~times}}.\]
The exponential function is easy to calculate:

\vskip +10 pt \noindent
{\bf Lemma.}
{\em
\noindent The power $ g^k $ can be calculated in at most $ 2 \log_2 k $ group 
operations.
}

\vskip +10 pt
\begin{Proof}{}
Let $ k=2^n + k_{n-1} 2^{n-1} + \cdots + k_1 2 + k_0 $ be the binary 
representation of $k. $ Then $ n \leq \log_2 (k), $  because $ 2^n \leq k < 
2^{n+1}. $ $ k $ can be written in the form $ k=2k' + k_0 $ with $ k'= 2^{n-1} + 
k_{n-1} 2^{n-2} + \cdots + k_1 $. Thus 
\[ g^k = g^{2k'+k_0}= (g^{k'})^2 g^{k_0} .\]
We therefore obtain $ g^k $ from $ g^{k'} $ by squaring and then multiplying by 
$ g $. The claim is thus proved by induction to $ n. $
\end{Proof}

\vskip +10 pt
\noindent {\bf Problem of the discrete logarithm}\index{Logarithm problem!discrete}
\vskip +10 pt
\begin{center}
\fbox{\parbox{12cm}{        
Let $ G $ by a finite group with the operation $ \circ. $ Let $ \alpha \in G $ 
and $ \beta \in H=\{ \alpha^{i}: i\geq 0\}. $ \\
We need to find a unique $ a \in {\rm N} $ with $ 0 \leq a \leq |H| -1 $ and $ \beta 
= \alpha^{a}. $ \\       
We define $ a $ as $ \log_\alpha (\beta). $
}}
\end{center}

@@

\noindent If $ |G| $ contains a ``dominant'' prime factor $ p ,$ then the 
complexity \index{Complexity} of the logarithm problem is approximately
\[ O(\sqrt{p}). \]

\noindent Therefore, if the logarithm problem is to be made difficult, the order of the 
group used $ G $ should have a large prime factor. In particular, if the 
discrete exponential function in the group $ \Z_p^{\ast} $ is to be a one way 
function, then $ p -1 $ must be a large prime factor. In this case a
generatlized ElGamal procedure can be defined (crypto
procedure~\ref{genelgamalproc}).
\begin{cryptoprocedure}
Let $ G $ be a finite group with operation $  \circ, $ and let $ \alpha \in G, $ 
so that the discrete logarithm in $ H =\{ \alpha^{i}: i \geq 0 \} $ is 
difficult, Let $ a $ with $   0 \leq a \leq |H| -1 $ and let $ \beta = 
\alpha^{a}. $\\
\textbf{Public key:}
\[ \alpha,\beta. \]
\textbf{Private key:}
\[a. \]
Let $ k \in \Z_{|H|} $ be a random number and $ x \in G $ be a plaintext. \\
\textbf{Encryption:}
\[ e_T(x,k)=(y_1,y_2), \]
where
\[ y_1=\alpha^k \]
and
\[ y_2 = x\circ \beta^k .\]
\textbf{Decryption:}
\[ d_T(y_1,y_2)= y_2\circ (y_1^{a})^{-1}  \]
\caption{Generalized ElGamal (based on the factorization problem)}
\label{genelgamalproc}
\end{cryptoprocedure}

\noindent \hyperlink{ellcurve}{Elliptic curves} {} provide useful groups for public key 
encryption procedures.

@@

%--------------------------------------------------------------------
\newpage
\chapter*{Web links}
% \section*{Web links}   %%% TODO: einheitlich section oder chapter für Web-Links
                         %%% TODO: einheitlich addcontentsline NACH WebLinks und NACH Bibliography
\addcontentsline{toc}{section}{Web links}
\begin{enumerate}
   \item \href{http://www.geocities.com/st_busygin/clipat.html}
              {http://www.geocities.com/st\_busygin/clipat.html }
\end{enumerate}
